Constraining gravitational wave amplitude birefringence with GWTC-3

### 2023-02-22 src/tex/ms.tex
Add tables

\begin{figure}
    \script{birefringence.py}
    \includegraphics[width=\columnwidth]{figures/birefringence.pdf}
    \caption{
        \emph{Illustration of amplitude birefringence.} The GR waveform for the $\ell=|m|=2$ mode of a nonprecessing BBH seen edge-on ($\cos\iota = 0$) is lineraly polarized and thus contains equal amounts of left- and right-handed modes for all frequencies (dotted, top).
        However, if spacetime were birefringent following Eq.~\protect\eqref{eq:waveform_modification}, the waveform observed on Earth would contain different fractions of the two circular modes, with higher frequencies affected more strongly (solid, top). 
        In the time domain, this manifests as a time-dependent amplification of the waveform, with a stronger effect at later times when the chirp reaches a higher instantaneous frequency (bottom).
        For this example, the black holes do not spin and have equal masses $m_1 = m_2 = 10\, M_\odot$, and we have chosen a luminosity distance $d_L = 400\, {\rm Mpc}$ and $\kappa = 0.6$.
        }
    \label{fig:birefringence}
\end{figure}

@@

For GW190521, we increase the maximum distance of the distance prior to $1.5\times$ the original value, as the birefringence effect enables the probability of a larger distance.
For GW190720, we decrease the analysis segment to 8 s from 16 s, in order to accomodate missing data near the edges of the 16 s segment in Virgo.

@@

If \ac{GR} is correct and there is no birefringence, we should find the observed $\kappa$ distribution to be consistent with a delta function at the origin ($\kappa_i = 0$ for all $i$, or $\mu=\sigma=0$); on the other hand, if spacetime is birefringent, we expect to find a delta function at some nonzero value ($\kappa_i = \kappa \neq 0$, or $\mu = \kappa$ and $\sigma=0$).
But this analysis also has the power to reveal unexpected physics or systematics in our measurements: if $\sigma$ is confidently found to be nonzero, this would imply that our set of measurements are statistically unlikely to originate from a unique $\kappa$ value.
This could signal further physics than is implied by Eq.~\eqref{eq:waveform_modification} \mi{which physics? @Will, any ideas?} or, more prosaically, that there are outliers in our measurements due to mismodeling, e.g., in the waveform approximant or the noise of the detector.

@@

Figure \ref{fig:violin_kappa} makes it clear that not all \acp{BBH} in GWTC-3 are equally informative about birefringence.
When considered individually, the events that best constrain $\kappa$ are \mi{THOMAS: list 3-5 events with the narrowest posterior---maybe also add a little table showing their median $\pm$ their 90\% CL, like $\tilde{\kappa}^{+x}_{-y}$ where $\tilde{\kappa}$ is the posterior median value, $x$ is the distance from the median to the 95\% percentile and $y$ is the distance to the 5\% percentile}

The event that shows the largest tension with $\kappa = 0$ is GW170818, for which $|\mu / \sigma| = \dots$ \mi{THOMAS: fill in these values}
Additionally, the GW200129 result shows features that could be indicative of contamination from non-Gaussian instrumental noise, so we exclude it from the collective analysis.
We discuss both these events in detail below.

Another set of events stand out in Fig.~\ref{fig:violin_kappa} due evident bimodality in the $\kappa$ posterior.
That is the case for GW170104, GW190413\_134308, GW190521, GW190805\_211137, and GW191105\_143521.
\mi{THOMAS: can you list the total masses of these events?}
We understand this effect to be a consequence of the narrow sets of frequencies that are visible for these heavy events, as we mentioned in Sec.~\ref{sec:inclination} and further demonstrate below.

@@

\begin{figure}
    \script{corner_GW150914.py}
    \includegraphics[width=\columnwidth]{figures/corner_GW150914.pdf}
    \caption{
        The posterior of $\kappa$, luminosity distance $d_L$ and $\cos{\iota}$ for GW150914.
        Colors in the plot represent the PE result with GR done by LVK without cosmological reweighing \citep{GWTC-2.1, GWTC-3}, the PE results done by us with both frequency independent and dependent birefringence respectively.
        2D contour plot shows $39.35\%$ and $90\%$ confidence level.
        Note that there is no posterior of $\kappa$ for the PE result from LVK, as the LVK PE is based on GR, which does not suggest GW amplitude birefringence.
        This plot shows that frequency-independent birefringence creates a degeneracy between $\kappa$ and $\iota$, while frequency-dependent birefringence can break it.
    }
    \label{fig:corner_GW150914}
\end{figure}

@@

\section{Illustration of birefringence}
\label{sec:birefringence_appendix}

>>>>

\begin{figure}
    \script{birefringence.py}
    \includegraphics[width=\columnwidth]{figures/birefringence.pdf}
    \caption{
        \emph{Illustration of amplitude birefringence.} The GR waveform for the $\ell=|m|=2$ mode of a nonprecessing BBH seen edge-on ($\cos\iota = 0$) is linearly polarized and thus contains equal amounts of left- and right-handed modes for all frequencies (dotted, top).
        However, if spacetime were birefringent following Eq.~\protect\eqref{eq:waveform_modification}, the waveform observed on Earth would contain different fractions of the two circular modes, with higher frequencies affected more strongly (solid, top). 
        In the time domain, this manifests as a time-dependent amplification of the waveform, with a stronger effect at later times when the chirp reaches a higher instantaneous frequency (bottom).
        For this example, the black holes do not spin and have equal masses $m_1 = m_2 = 10\, M_\odot$, and we have chosen a luminosity distance $d_L = 400\, {\rm Mpc}$ and $\kappa = 0.6$.
        }
    \label{fig:birefringence}
\end{figure}

@@

For GW190521, we increase the maximum distance of the distance prior to $1.5\times$ the original value, as the birefringence effect enables the probability of a larger distance.
For GW190720, we decrease the analysis segment to 8 s from 16 s, in order to accommodate missing data near the edges of the 16 s segment in Virgo.

@@

If \ac{GR} is correct and there is no birefringence, we should find the observed $\kappa$ distribution to be consistent with a delta function at the origin ($\kappa_i = 0$ for all $i$, or $\mu=\sigma=0$); on the other hand, if spacetime is birefringent, we expect to find a delta function at some nonzero value ($\kappa_i = \kappa \neq 0$, or $\mu = \kappa$ and $\sigma=0$).
But this analysis also has the power to reveal unexpected physics or systematics in our measurements: if $\sigma$ is confidently found to be nonzero, this would imply that our set of measurements is statistically unlikely to originate from a unique $\kappa$ value.
This could signal further physics than is implied by Eq.~\eqref{eq:waveform_modification} \mi{which physics? @Will, any ideas?} or, more prosaically, that there are outliers in our measurements due to mismodeling, e.g., in the waveform approximant or the noise of the detector.

@@

Figure \ref{fig:violin_kappa} makes it clear that not all \acp{BBH} in GWTC-3 are equally informative about birefringence.
When considered individually, the events that best constrain $\kappa$ are listed in Table~\ref{tab:best_events}.
\begin{table}
    \begin{ruledtabular}
        \begin{tabular}{ll}
            Event & $\tilde{\kappa}$\\
            \hline
            \variable{output/best_events.txt}
        \end{tabular}
    \end{ruledtabular}
    \caption{
        Events that best constrain $\kappa$.
        $\tilde{\kappa}$ is the posterior median of $\kappa$.
    }
    \label{tab:best_events}
\end{table}

The event that shows the largest tension with $\kappa = 0$ is GW170818, for which $|\mu / \sigma| =$ \variable{output/GW170818_constraint.txt}.
Additionally, the GW200129 result shows features that could be indicative of contamination from non-Gaussian instrumental noise, so we exclude it from the collective analysis.
We discuss both of these events in detail below.

Another set of events stands out in Fig.~\ref{fig:violin_kappa} due to evident bimodality in the $\kappa$ posterior.
That is the case for GW170104, GW190413\_134308, GW190521, GW190805\_211137, and GW191105\_143521.
\begin{table}\
    \begin{ruledtabular}
        \begin{tabular}{ll}
            Event & Total Mass ($(M_{\odot})$)\\
            \hline
            \variable{output/bimodal_events_mass.txt}
        \end{tabular}
    \end{ruledtabular}
    \caption{
        Events that show bimodality in the $\kappa$ posterior.
        The total mass of the binary is given in solar masses.
    }
    \label{tab:bimodal_events_mass}
\end{table}
We understand this effect to be a consequence of the narrow sets of frequencies that are visible for these heavy events, as we mentioned in Sec.~\ref{sec:inclination} and further demonstrate below.

@@

\begin{figure}
    \script{corner_GW150914.py}
    \includegraphics[width=\columnwidth]{figures/corner_GW150914.pdf}
    \caption{
        The posterior of $\kappa$, luminosity distance $d_L$ and $\cos{\iota}$ for GW150914.
        Colors in the plot represent the PE result with GR done by LVK without cosmological reweighing \citep{GWTC-2.1, GWTC-3}, the PE results done by us with both frequency-independent and dependent birefringence respectively.
        2D contour plot shows $39.35\%$ and $90\%$ confidence level.
        Note that there is no posterior of $\kappa$ for the PE result from LVK, as the LVK PE is based on GR, which does not suggest GW amplitude birefringence.
        This plot shows that frequency-independent birefringence creates a degeneracy between $\kappa$ and $\iota$, while frequency-dependent birefringence can break it.
    }
    \label{fig:corner_GW150914}
\end{figure}

@@
